---
title: "forestfire_analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import packages
```{r}
library(ggplot2)
library(plotly)
library(pheatmap)
library(corrplot)
```

## Load Data
```{r}
fires <- read.csv("forestfires.csv", header = TRUE, sep = ",")

# View(fires)
```

## Part 1: Descriptive Analysis

### Distribution of response variable
```{r}
summary(fires$area)
```

```{r}
# Boxplots
ggplot(fires, aes(y = area)) + geom_boxplot()
ggplot(fires, aes(y = area)) + geom_boxplot() + ylim(0,125)
```

# Histogram
```{r}
ggplot(fires, aes(x = area)) + geom_histogram()
ggplot(fires, aes(x = area)) + geom_histogram() + xlim(-1, 100) + ylim(0, 100)

```
### Distribution of categorical explanatory variable (month and region)


## Part 2: One-sample statistical inference for response variable ('area')

```{r}

```


## Part 3: Two-sample statistical inference for categorical variable 

## Part 4: Two-sample statistical inference

## Part 5: Which explanatory variables are the best predictors for 'area'?
### Correlation plot of numeric explanatory variables

```{r}
# Paul, edit as you wish. I'm not sure which of the columns you will want to obtain correlations for in the 'select' statement. 
corr_matrix <- fires %>%
                  select(FFMC, DMC, DC, ISI, temp, RH, wind, rain, area) %>%
                  cor
corr_matrix
corrplot(corr_matrix)
```

